PineAPPL: combining EW and QCD corrections\\ for fast evaluation of LHC processes

### 2020-10-14 paper/answer_to_the_editor.txt

Dear Editor,

We would like to thank the referee for carefully reading the manuscript and
making useful comments and suggestions.

In the following our answers to the issues raised by the referee:

- Issues 1, 3, 4, 5, 6, 8, 9, and 11 have all been addressed as suggested or
  corrected otherwise.
- Issue 2: The decomposition of the fixed-order expansion written down in eq.
  (2.1) holds to all orders in both the strong and electroweak coupling, and
  therefore the mentioned sentence is correct—it only concerns PineAPPL. The
  Monte Carlo that we used is tied to NLO, of course.
- Issue 7:
- Issue 10:
- Issue 12: Although we agree with the referee that the mentioned tables are
  interesting only to a few readers, the main text refers to it in a few
  places and therefore removing it from the main document would require bigger
  changes. We see no harm in leaving the tables in the document and therefore
  decided to keep them.

Note that all changes have been highlighted in red in the text.

With these changes, we hope that the current version will meet the criteria of
publication of JHEP.

Best regards,
The authors

### 2020-10-14 paper/answer_to_the_editor.txt

Dear Editor,

We would like to thank the referee for carefully reading the manuscript and
making useful comments and suggestions.

In the following our answers to the issues raised by the referee:

- Issues 1, 3, 4, 5, 6, 8, 9, and 11 have all been addressed as suggested or
  corrected otherwise.
- Issue 2: The decomposition of the fixed-order expansion written down in eq.
  (2.1) holds to all orders in both the strong and electroweak coupling, and
  therefore the mentioned sentence is correct—it only concerns PineAPPL. The
  Monte Carlo that we used is tied to NLO, of course.
- Issue 7:
- Issue 10: A brief discussion on this matter has been added at the beginning
  of section 3. The reference suggested (Buccioni:2020cfi), is cited in section 
  3.2.1, in the discussion relevant for Drell-Yan.
- Issue 12: Although we agree with the referee that the mentioned tables are
  interesting only to a few readers, the main text refers to it in a few
  places and therefore removing it from the main document would require bigger
  changes. We see no harm in leaving the tables in the document and therefore
  decided to keep them.

### 2020-10-16 paper/app-pineappl.tex
Reintroduces multiplication signs
- with correct encoding this should work now
Fix errors found by `lacheck`

\paragraph{Channels and Luminosity function.}
Sometimes it is useful to know which partons contribute the most and by how much.
This is what the subcommand \texttt{channels} shows:
\begin{verbatim}
 $ pineappl channels ATLASZHIGHMASS49FB.pineappl \
 > NNPDF31_nlo_as_0118_luxqed --limit 5
bin xmin xmax lumi  size
---+----+----+----+------+---+------+---+------+---+------+---+-----
  0  116  130  #15 27.42%  #0 27.42%  #5 24.52% #20 24.51% #30 1.30%
  1  130  150  #15 28.83%  #0 28.83%  #5 21.78% #20 21.78% #30 1.86%
  2  150  170  #15 29.87%  #0 29.82%  #5 19.63% #20 19.62% #30 2.32%
  3  170  190  #15 30.54%  #0 30.53% #20 18.17%  #5 18.13% #30 2.58%
  4  190  210   #0 31.06% #15 31.04%  #5 17.07% #20 17.04% #30 2.75%
  5  210  230   #0 31.46% #15 31.38% #20 16.27%  #5 16.24% #30 2.86%
  6  230  250   #0 31.73% #15 31.64% #20 15.66%  #5 15.62% #30 2.91%
  7  250  300   #0 32.07% #15 32.02%  #5 14.91% #20 14.86% #30 2.98%
  8  300  400   #0 32.62% #15 32.56%  #5 13.83% #20 13.81% #30 2.98%
  9  400  500  #15 33.14%  #0 33.12% #20 12.89%  #5 12.87% #30 2.93%
 10  500  700   #0 33.71% #15 33.31% #20 12.28%  #5 12.26% #30 2.93%
 11  700 1000  #15 34.08%  #0 33.71% #20 11.59%  #5 11.51% #30 3.06%
 12 1000 1500   #0 33.95% #15 33.88% #20 11.12%  #5 10.88% #30 3.56%
\end{verbatim}
The first three columns are known from \texttt{convolute}.
The next columns (the switch \texttt{-{}-limit 5} limits the output to five columns) show first the channel index and then the relative size of the corresponding contribution.
Since the contribution of a partonic channel can be negative, the columns are sorted ignoring the sign of the contribution.
The first line shows that for bin 0, i.e.\ for the range $\SI{116}{\giga\electronvolt} < M_{\ell \bar{\ell}} < \SI{130}{\giga\electronvolt}$, the cross section is dominated by partonic channel \texttt{\#15} (\SI{27.42}{\percent}), following by partonic channel \texttt{\#0} with same size, then channel \texttt{\#5}, etc.
The meaning of the channel numbers is given by using the subcommand \texttt{luminosity} (only an excerpt is shown):
\begin{verbatim}
id    entry
--+------------+------------
0  1 x ( 2, -2) 1 x ( 4, -4)
5  1 x ( 1, -1) 1 x ( 3, -3)
15 1 x (-4,  4) 1 x (-2,  2)
20 1 x (-3,  3) 1 x (-1,  1)
30 1 x (22, 22)
\end{verbatim}
This shows that channel \texttt{\#0} represents the up-type quark--anti-quark contributions (shown with PDG id 2 and 4 for up and charm quarks, which have the same matrix elements), channel \texttt{\#15} is the same channel with its initial states transposed, channels \texttt{\#5} and \texttt{\#20} are the down-type quark--anti-quark channels, and channel \texttt{\#30} is the photon--photon channel.
The size of the remaining channels is smaller than the photon--photon channel.
The factors \texttt{1} are not important here, but in general they can contain CKM values and charge factors that, if kept in the squared matrix elements, would not allow for sharing a single matrix element for different quark flavours and therefore slow down the calculation.
A complete list of all channels and of their contribution to the cross section for all of the processes discussed in section~\ref{sec:results} is collected in appendix~\ref{app:lumis}.

@@

The following listing shows how to setup \textsc{PineAPPL} using its C interface in a simple Monte Carlo integrator for calculating the double-photon contribution to Drell--Yan lepton-pair production at the LHC.
All \textsc{PineAPPL} functions have the prefix \texttt{pineappl\_}.
The full example together with a makefile can be found at \url{https://github.com/N3PDF/pineappl/tree/master/examples/capi-dy-aa}.
The documentation of the C API can be found at \url{https://docs.rs/pineappl_capi}.

>>>>

\paragraph{Channels and Luminosity function.}
Sometimes it is useful to know which partons contribute the most and by how much.
This is what the subcommand \texttt{channels} shows:
\begin{verbatim}
 $ pineappl channels ATLASZHIGHMASS49FB.pineappl \
 > NNPDF31_nlo_as_0118_luxqed --limit 5
bin xmin xmax lumi  size
---+----+----+----+------+---+------+---+------+---+------+---+-----
  0  116  130  #15 27.42%  #0 27.42%  #5 24.52% #20 24.51% #30 1.30%
  1  130  150  #15 28.83%  #0 28.83%  #5 21.78% #20 21.78% #30 1.86%
  2  150  170  #15 29.87%  #0 29.82%  #5 19.63% #20 19.62% #30 2.32%
  3  170  190  #15 30.54%  #0 30.53% #20 18.17%  #5 18.13% #30 2.58%
  4  190  210   #0 31.06% #15 31.04%  #5 17.07% #20 17.04% #30 2.75%
  5  210  230   #0 31.46% #15 31.38% #20 16.27%  #5 16.24% #30 2.86%
  6  230  250   #0 31.73% #15 31.64% #20 15.66%  #5 15.62% #30 2.91%
  7  250  300   #0 32.07% #15 32.02%  #5 14.91% #20 14.86% #30 2.98%
  8  300  400   #0 32.62% #15 32.56%  #5 13.83% #20 13.81% #30 2.98%
  9  400  500  #15 33.14%  #0 33.12% #20 12.89%  #5 12.87% #30 2.93%
 10  500  700   #0 33.71% #15 33.31% #20 12.28%  #5 12.26% #30 2.93%
 11  700 1000  #15 34.08%  #0 33.71% #20 11.59%  #5 11.51% #30 3.06%
 12 1000 1500   #0 33.95% #15 33.88% #20 11.12%  #5 10.88% #30 3.56%
\end{verbatim}
The first three columns are known from \texttt{convolute}.
The next columns (the switch \texttt{-{}-limit 5} limits the output to five columns) show first the channel index and then the relative size of the corresponding contribution.
Since the contribution of a partonic channel can be negative, the columns are sorted ignoring the sign of the contribution.
The first line shows that for bin 0, i.e.\ for the range $\SI{116}{\giga\electronvolt} < M_{\ell \bar{\ell}} < \SI{130}{\giga\electronvolt}$, the cross section is dominated by partonic channel \texttt{\#15} (\SI{27.42}{\percent}), following by partonic channel \texttt{\#0} with same size, then channel \texttt{\#5}, etc.
The meaning of the channel numbers is given by using the subcommand \texttt{luminosity} (only an excerpt is shown):
\begin{verbatim}
id    entry
--+------------+------------
0  1 × ( 2, -2) 1 × ( 4, -4)
5  1 × ( 1, -1) 1 × ( 3, -3)
15 1 × (-4,  4) 1 × (-2,  2)
20 1 × (-3,  3) 1 × (-1,  1)
30 1 × (22, 22)
\end{verbatim}
This shows that channel \texttt{\#0} represents the up-type quark--anti-quark contributions (shown with PDG id 2 and 4 for up and charm quarks, which have the same matrix elements), channel \texttt{\#15} is the same channel with its initial states transposed, channels \texttt{\#5} and \texttt{\#20} are the down-type quark--anti-quark channels, and channel \texttt{\#30} is the photon--photon channel.
The size of the remaining channels is smaller than the photon--photon channel.
The factors \texttt{1} are not important here, but in general they can contain CKM values and charge factors that, if kept in the squared matrix elements, would not allow for sharing a single matrix element for different quark flavours and therefore slow down the calculation.
A complete list of all channels and of their contribution to the cross section for all of the processes discussed in section~\ref{sec:results} is collected in appendix~\ref{app:lumis}.

@@

The following listing shows how to setup \textsc{PineAPPL} using its C interface in a simple Monte Carlo integrator for calculating the double-photon contribution to Drell--Yan lepton-pair production at the LHC\@.
All \textsc{PineAPPL} functions have the prefix \texttt{pineappl\_}.
The full example together with a makefile can be found at \url{https://github.com/N3PDF/pineappl/tree/master/examples/capi-dy-aa}.
The documentation of the C API can be found at \url{https://docs.rs/pineappl_capi}.

### 2020-10-16 paper/sec-pineappl.tex
Fix errors found by `lacheck`

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u(y(x_1)),u(y(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula,
\begin{equation}
a_{i,j} \leftarrow a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2) \text{,}
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions,
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k} \text{,}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remap
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid point index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part of $u(y)$ gives the relative location between the nearest grid points.

@@

Note that all initial states have to be taken into account that lead to the same final state.
This includes the photon--photon initial state, which appears already at LO.
In the corresponding Feynman diagrams all particles are colourless, so that the photon--photon initiated contributions only receive EW corrections.
The EW corrections also introduce quark--photon contributions, in analogy of QCD corrections introducing quark--gluon contributions.

>>>>

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in the $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u(y(x_1)),u(y(x_2)))$, given by the MC\@.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula,
\begin{equation}
a_{i,j} \leftarrow a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2) \text{,}
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions,
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k} \text{,}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting from the smallest index in the square, $k_i$ and $k_j$.
Finally, we remap
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid point index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part of $u(y)$ gives the relative location between the nearest grid points.

@@

Note that all initial states have to be taken into account that lead to the same final state.
This includes the photon--photon initial state, which appears already at LO\@.
In the corresponding Feynman diagrams all particles are colourless, so that the photon--photon initiated contributions only receive EW corrections.
The EW corrections also introduce quark--photon contributions, in analogy of QCD corrections introducing quark--gluon contributions.

### 2020-10-16 paper/sec-results.tex
Fix fixes
Fix errors found by `lacheck`

In this section we demonstrate the capabilities of the \textsc{PineAPPL}
library by interfacing it to \textsc{mg5\_aMC}, and by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a common set of LHC processes in which EW corrections may sizeably
affect the accuracy of the theoretical predictions.
In order to consider realistic kinematics for these
processes, we rely on a representative set of LHC measurements. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for the specific experimental setups. We describe
the settings employed for the computations, the corresponding results for
each process, and possible implications for the determination of PDFs. It is worth to mention
that, in those phase-space regions where both NLO QCD and EW corrections are sizeable, the NNLO
correction at $\mathcal O (\alphas \alpha)$ is also expected to be important, and has either to
be included, when it is known exactly or at least in some approximate way, or to be accounted for 
as an extra uncertainty. We leave this aspect to future studies

@@

We present these comparisons in
figures~\ref{fig:atlaszhighmass49fb}, \ref{fig:cmsdy2d11_bins3456}, \ref{fig:atlastop}, and \ref{fig:cmsZ13TeV}.
The format of the plots is the same across all figures. The first panel
displays the relative difference (in per mille) between the {\sc PineAPPL} and
the MC results for the central scale choice and upper/lower edges of the
scale-uncertainty envelope, for
theories accurate to both NLO QCD and NLO QCD+EW. The following three panels
present the theoretical predictions, accurate to either NLO QCD or NLO QCD+EW,
always normalised to the former; on top of the theoretical predictions, the
PDF uncertainty, the scale uncertainty and the Monte Carlo uncertainty are
displayed in turn. The relative uncertainty of the experimental data is
also shown for comparison. We shall now discuss the results for each
process and data set.

@@

%-------------------------------------------------------------------------------
\begin{figure}[!t]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/pineappl_ATLASZHIGHMASS49FB}\\
    \caption{Validation and test of the \textsc{PineAPPL} grid for the ATLAS
      Drell--Yan lepton pair measurement in the high-mass region at
      a centre-of-mass energy of \SI{7}{\tera\electronvolt}~\cite{Aad:2013iua}. The first panel
      displays the relative difference (in per mille) between the {\sc PineAPPL}
      and the MC results for the central, upper and lower scale choices,
      for theories accurate to both NLO QCD and NLO QCD+EW. The second, third
      and fourth panels present the theoretical predictions, accurate to either
      NLO QCD or NLO QCD+EW, always normalised to the former; on top of the
      theoretical predictions, the PDF uncertainty, the scale uncertainty and
      the Monte Carlo uncertainty are displayed in turn. The relative
      uncertainty of the experimental data is also shown for comparison.}
    \label{fig:atlaszhighmass49fb}
\end{figure}
%-------------------------------------------------------------------------------

>>>>

In this section we demonstrate the capabilities of the \textsc{PineAPPL}
library by interfacing it to \textsc{mg5\_aMC}, and by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a common set of LHC processes in which EW corrections may sizeably
affect the accuracy of the theoretical predictions.
In order to consider realistic kinematics for these
processes, we rely on a representative set of LHC measurements. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for the specific experimental setups. We describe
the settings employed for the computations, the corresponding results for
each process, and possible implications for the determination of PDFs. It is worth to mention
that, in those phase-space regions where both NLO QCD and EW corrections are sizeable, the NNLO
correction at $\mathcal O (\alphas \alpha)$ is also expected to be important, and has either to
be included, when it is known exactly or at least in some approximate way, or to be accounted for 
as an extra uncertainty. We leave this aspect to future studies.

@@

We present these comparisons in
figures~\ref{fig:atlaszhighmass49fb}, \ref{fig:cmsdy2d11_bins3456}, \ref{fig:atlastop}, and \ref{fig:cmsZ13TeV}.
The format of the plots is the same across all figures. The first panel
displays the relative difference (in per mille) between the {\sc PineAPPL} and
the MC results for the central scale choice and upper/lower edges of the
scale-uncertainty envelope, for
theories accurate to both NLO QCD and NLO QCD+EW\@. The following three panels
present the theoretical predictions, accurate to either NLO QCD or NLO QCD+EW,
always normalised to the former; on top of the theoretical predictions, the
PDF uncertainty, the scale uncertainty and the Monte Carlo uncertainty are
displayed in turn. The relative uncertainty of the experimental data is
also shown for comparison. We shall now discuss the results for each
process and data set.

@@

%-------------------------------------------------------------------------------
\begin{figure}[!t]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/pineappl_ATLASZHIGHMASS49FB}\\
    \caption{Validation and test of the \textsc{PineAPPL} grid for the ATLAS
      Drell--Yan lepton pair measurement in the high-mass region at
      a centre-of-mass energy of \SI{7}{\tera\electronvolt}~\cite{Aad:2013iua}. The first panel
      displays the relative difference (in per mille) between the {\sc PineAPPL}
      and the MC results for the central, upper and lower scale choices,
      for theories accurate to both NLO QCD and NLO QCD+EW\@. The second, third
      and fourth panels present the theoretical predictions, accurate to either
      NLO QCD or NLO QCD+EW, always normalised to the former; on top of the
      theoretical predictions, the PDF uncertainty, the scale uncertainty and
      the Monte Carlo uncertainty are displayed in turn. The relative
      uncertainty of the experimental data is also shown for comparison.}
    \label{fig:atlaszhighmass49fb}
\end{figure}
%-------------------------------------------------------------------------------
